---
title: "IRS 990 — Homeless Services Compensation Explorer"
output: html_notebook
---

## Setup

Install packages once if you don't have them:
```{r install, eval=FALSE}
install.packages(c("DBI", "RPostgres", "tidyverse", "scales"))
```

Load and connect:
```{r setup, message=FALSE}
library(DBI)
library(RPostgres)
library(tidyverse)
library(scales)

con <- dbConnect(
  RPostgres::Postgres(),
  host     = "localhost",
  port     = 5433,       # Docker is on 5433, not 5432
  dbname   = "irs990",
  user     = "irs990",
  password = "irs990"
)
```

---

## How much data do we have?

```{r counts}
dbGetQuery(con, "
  SELECT
    tax_year,
    COUNT(DISTINCT ein)    AS orgs,
    COUNT(DISTINCT object_id) AS filings,
    COUNT(*)               AS comp_records
  FROM schedule_j_compensation
  GROUP BY tax_year
  ORDER BY tax_year
")
```

---

## Top cities by number of orgs

```{r top_cities}
dbGetQuery(con, "
  SELECT
    city,
    state,
    COUNT(DISTINCT ein) AS orgs
  FROM schedule_j_compensation
  WHERE city IS NOT NULL
  GROUP BY city, state
  ORDER BY orgs DESC
  LIMIT 25
")
```

---

## Total executive compensation by city/state (all years)

```{r comp_by_city}
comp_by_city <- dbGetQuery(con, "
  SELECT
    city,
    state,
    SUM(total_comp_filing_org)          AS total_comp,
    COUNT(DISTINCT ein)                 AS orgs,
    COUNT(*)                            AS exec_records,
    AVG(total_comp_filing_org)          AS avg_exec_comp
  FROM schedule_j_compensation
  WHERE total_comp_filing_org > 0
    AND city IS NOT NULL
  GROUP BY city, state
  ORDER BY total_comp DESC
  LIMIT 30
")

comp_by_city %>%
  mutate(
    total_comp    = dollar(as.numeric(total_comp)),
    avg_exec_comp = dollar(round(as.numeric(avg_exec_comp)))
  )
```

---

## Compensation trend over time — top 10 cities

```{r trend}
trend <- dbGetQuery(con, "
  SELECT
    city,
    state,
    tax_year,
    SUM(total_comp_filing_org) AS total_comp,
    COUNT(DISTINCT ein)        AS orgs
  FROM schedule_j_compensation
  WHERE total_comp_filing_org > 0
    AND city IN (
      SELECT city FROM schedule_j_compensation
      WHERE city IS NOT NULL
      GROUP BY city
      ORDER BY COUNT(DISTINCT ein) DESC
      LIMIT 10
    )
  GROUP BY city, state, tax_year
  ORDER BY city, tax_year
")

ggplot(trend %>% mutate(total_comp = as.numeric(total_comp)),
       aes(x = tax_year, y = total_comp, color = city)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_y_continuous(labels = dollar_format(scale = 1e-6, suffix = "M")) +
  labs(
    title = "Total Exec Compensation — Homeless Services Orgs",
    subtitle = "Top 10 cities by org count",
    x = "Tax Year",
    y = "Total Compensation",
    color = "City"
  ) +
  theme_minimal()
```

---

## Highest-paid executives across all orgs

```{r top_execs}
dbGetQuery(con, "
  SELECT
    person_name,
    title,
    company_name,
    city,
    state,
    tax_year,
    total_comp_filing_org   AS total_comp,
    base_compensation,
    bonus
  FROM schedule_j_compensation
  WHERE total_comp_filing_org > 0
  ORDER BY total_comp_filing_org DESC
  LIMIT 25
") %>%
  mutate(
    total_comp        = dollar(as.numeric(total_comp)),
    base_compensation = dollar(as.numeric(base_compensation)),
    bonus             = dollar(as.numeric(bonus))
  )
```

---

## Org-level summary — average CEO pay vs number of filings

```{r org_summary}
dbGetQuery(con, "
  SELECT
    company_name,
    ein,
    city,
    state,
    ntee_code,
    COUNT(DISTINCT tax_year)          AS years_filed,
    AVG(total_comp_filing_org)        AS avg_total_comp,
    MAX(total_comp_filing_org)        AS max_total_comp
  FROM schedule_j_compensation
  WHERE total_comp_filing_org > 0
  GROUP BY company_name, ein, city, state, ntee_code
  ORDER BY avg_total_comp DESC
  LIMIT 30
") %>%
  mutate(
    avg_total_comp = dollar(round(as.numeric(avg_total_comp))),
    max_total_comp = dollar(as.numeric(max_total_comp))
  )
```

---

## Portland vs Los Angeles — side-by-side

```{r pdx_vs_la}
dbGetQuery(con, "
  SELECT
    city,
    state,
    tax_year,
    COUNT(DISTINCT ein)            AS orgs,
    SUM(total_comp_filing_org)     AS total_exec_comp,
    AVG(total_comp_filing_org)     AS avg_exec_comp,
    MAX(total_comp_filing_org)     AS max_exec_comp
  FROM schedule_j_compensation
  WHERE (city ILIKE '%portland%'  AND state = 'OR')
     OR (city ILIKE '%los angeles%' AND state = 'CA')
  GROUP BY city, state, tax_year
  ORDER BY city, tax_year
")
```

---

## NTEE code breakdown — what types of orgs are in the data?

```{r ntee_breakdown}
dbGetQuery(con, "
  SELECT
    ntee_code,
    COUNT(DISTINCT ein) AS orgs,
    SUM(total_comp_filing_org) AS total_comp
  FROM schedule_j_compensation
  WHERE ntee_code IS NOT NULL
  GROUP BY ntee_code
  ORDER BY orgs DESC
")
```

---

## Cleanup

```{r disconnect}
dbDisconnect(con)
```
